---
title: "Touchstone"
output: github_document
bibliography: tstone_readme.bib
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd. -->

```{r setup, include=FALSE}
devtools::load_all(".", quiet = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.width = 9,
  fig.height = 5.5,
  out.width = "100%",
  message = FALSE,
  warning = FALSE
)
options(width = 100)
```

## Crosslinking mass-spectrometry rescoring and reporting

Touchstone is an R package for rescoring and classifying crosslinking
mass-spectrometry (CLMS/XL-MS) results from
[Protein Prospector](https://prospector.ucsf.edu/prospector/mshome.htm).
It builds a support-vector-machine score from Search Compare output, estimates
separate intra- and interprotein false-discovery-rate (FDR) thresholds, and
prepares results at several proteomics summarization levels.

The current workflow supports:

- crosslinked spectrum matches (`"csm"`);
- unique residue pairs (`"urp"`);
- peptide pairs (`"peptide-pair"`);
- protein pairs (`"protein-pair"`); and
- user-defined module pairs (`"module-pair"`).

Model fitting, statistical classification, and optional evidence-quality
polishing are kept separate. Linear SVMs are the conservative default, and the
recommended score averages three reproducible cross-fitted estimates. The
Score.Diff training prefilter, feature profile, and SVM cost can be selected
automatically or fixed for a prespecified analysis.

Touchstone was developed for Protein Prospector CLMS workflows [@trnka_matching_2014].
It currently assumes familiarity with R and with the decoy-search design used
to generate the Prospector results.

## Installation

Touchstone is currently installed from GitHub:

```{r installation, eval=FALSE}
install.packages("remotes")
remotes::install_github("mtrnka/Touchstone")
library(touchstone)
```

The example below is included with the package and can be run in a fresh R
session after installation.

## Worked example: DSSO-crosslinked rabbit 80S ribosome

The example is a stepped-HCD MS2 analysis of rabbit 80S ribosomes produced in a
reticulocyte cell-free expression system and crosslinked with DSSO
[@kao_development_2011]. The Search Compare table contains eight LC-MS runs
covering four size-exclusion-chromatography fractions and two replicate series.
The search database contains 77 ribosomal proteins plus randomized decoy
sequences that are ten times longer than their corresponding targets.

### 1. Read the Search Compare output

The decoy scaling factor must match the database used for the Prospector
search. An incorrect value changes every FDR calculation. This example uses a
10-fold decoy database:

```{r read-ribosome}
library(touchstone)

setDecoyScalingFactor(10)

ribo <- readProspectorXLOutput(
  touchstone_example("M6.sthcd_scout.txt"),
  minPepLen = 4,
  minIons = 0
)

ribo_target_csms <- removeDecoys(ribo)

c(
  CSM_candidates = nrow(ribo),
  target_accessions = length(unique(c(
    ribo_target_csms$Acc.1,
    ribo_target_csms$Acc.2
  )))
)
```

`minIons = 0` deliberately avoids filtering product-ion evidence during import.
The evidence-quality rule is applied later, after model fitting. This keeps the
statistical training decision distinct from the reporting policy.

The recommended Prospector Search Compare parameter template can be located
with:

```{r search-compare-template}
search_compare_template <- touchstone_example("tstoneMS2.4.json")
basename(search_compare_template)
```

Some evidence-polishing features depend on optional MS-Product-derived columns.
Touchstone continues without a requested optional field and records that the
corresponding polishing rule was unavailable.

### 2. Train the crosslink score

`trainCrosslinkScore()` chooses a complexity-appropriate feature profile,
optionally selects a training-only Score.Diff threshold, evaluates the requested
hyperparameters once, and then averages three fits of the selected model. All
input CSMs are scored, including rows excluded from the training subset.

For a completely automatic analysis, use:

```{r automatic-training-call, eval=FALSE}
ribo_training <- trainCrosslinkScore(ribo, targetER = 0.01)
```

The fully automatic run on this bundled dataset selects a Score.Diff prefilter
of `0` and linear cost `0.001`. The executable README pins those validated
choices to avoid repeating the exploratory prefilter and cost grids every time
the documentation is rebuilt:

```{r train-ribosome}
ribo_training <- trainCrosslinkScore(
  ribo,
  targetER = 0.01,
  scoreDiffPrefilter = 0,
  cost_values = 0.001
)

ribo_training
```

The printed result reports the chosen complexity profile, features, prefilter,
decoy scaling factor, ensemble size, and candidate audit. The complete audit is
available as `ribo_training$candidates`, and the shared settings are in
`ribo_training$settings`.

### 3. Inspect score behavior

The trained score should retain a sensible within-class relationship with the
original Prospector Score.Diff. This diagnostic is particularly important when
testing flexible radial kernels or new features:

```{r score-correlation}
plotScoreCorrelation(
  ribo_training,
  alpha = 0.18,
  pointSize = 0.45
)
```

The recommended linear model is stored in `ribo_training$recommended`. Radial
models are not trained unless explicitly requested with
`kernels = c("linear", "radial")`.

### 4. Prepare results at the URP level

Training produces one CSM-level score. Reporting thresholds are estimated
separately at each summarization level. Here the complete scored CSM table is
summarized as unique residue pairs:

```{r prepare-urps}
ribo_urp <- prepareCrosslinkResults(
  ribo_training,
  summarizationLevel = "urp"
)

ribo_urp$thresholds
```

The prepared object retains the complete target-and-decoy URP table, its CSM
source, thresholds, settings, and classification summaries. It can therefore be
plotted before classification:

```{r plot-urp-fdr}
fdrPlots(
  ribo_urp,
  xLimits = c(-1.5, 2.5),
  title = "Rabbit 80S ribosome"
)
```

The interprotein FDR curve is modeled because sparse decoys can make the raw
empirical curve jagged. The requested FDR is an estimate, not a requirement
that every raw count-based diagnostic fall below a hard maximum.

### 5. Classify and polish the results

`classifyCrosslinkResults()` applies the evidence policy to the scored CSMs,
re-estimates the reporting-level thresholds, and recalculates `numCSM` and
`numURP` from evidence that passes the final policy. By default, each peptide
must have at least three distinct annotated product-ion cleavage positions when
those annotations are available:

```{r classify-urps}
ribo_report <- classifyCrosslinkResults(ribo_urp)

ribo_report$polishingAudit

ribo_target_summary <- ribo_report$data |>
  dplyr::filter(.data$Decoy == "Target") |>
  dplyr::count(.data$xlinkClass, name = "target_URPs")

ribo_target_summary
```

This analysis reports `r ribo_target_summary$target_URPs[ribo_target_summary$xlinkClass == "interProtein"]`
interprotein and
`r ribo_target_summary$target_URPs[ribo_target_summary$xlinkClass == "intraProtein"]`
intraprotein target URPs at the modeled 1% thresholds after polishing.

The final target-only table is obtained without storing another permanent copy
inside the result object:

```{r target-only-table}
ribo_targets <- removeDecoys(ribo_report$data)
ribo_targets[1:5, c(
  "xlinkedResPair", "xlinkClass", "SVM.score", "numCSM"
)]
```

More stringent policies can be requested explicitly. For example:

```{r stronger-polishing, eval=FALSE}
ribo_stringent <- classifyCrosslinkResults(
  ribo_urp,
  polishing = list(
    minIons = 4,
    minPepLen = 5,
    minLadderCoverage = 0.25
  )
)
```

If product-ion annotations are absent, `fallbackMinScoreDiff` can be supplied
alongside `minIons`; Touchstone records whether the ion rule, fallback, or
neither was applied.

### 6. Change the reporting level

The SVM is not retrained for each reporting level. The same scored CSM table can
be prepared and classified as peptide pairs or protein pairs:

```{r other-levels, eval=FALSE}
ribo_peptide_pairs <- prepareCrosslinkResults(
  ribo_training,
  summarizationLevel = "peptide-pair"
) |>
  classifyCrosslinkResults()

ribo_protein_pairs <- prepareCrosslinkResults(
  ribo_training,
  summarizationLevel = "protein-pair"
) |>
  classifyCrosslinkResults()
```

This separation is important: the CSM score is shared, but the FDR threshold
and supporting-evidence counts belong to the requested reporting level.

### 7. Optional structural annotation and export

The bundled module file maps ribosomal proteins to the 40S and 60S subunits and
to the rabbit 80S structure
[PDB 6HCJ](https://www.rcsb.org/structure/6HCJ). Structural annotation may
download coordinate files, so it is not executed while building this README:

```{r structural-validation, eval=FALSE}
ribo_structural <- processModuleFile(
  ribo_report$data,
  touchstone_example("rRibo_modfile_uniprot.txt")
)

distancePlot2(
  removeDecoys(ribo_structural),
  threshold = 35
)
```

Results can be formatted for downstream inspection, including MS-Viewer:

```{r export, eval=FALSE}
msviewer_table <- formatXLTable(ribo_targets, msviewer = TRUE)
readr::write_tsv(msviewer_table, "ribosome_urp_msviewer.txt")
```

MS-Viewer additionally requires the corresponding peak list; for large raw
files, a filtered peak list containing only reported spectra is usually more
practical.

## Scope and development status

The stabilized public path is currently binary-crosslink MS2 analysis from
Protein Prospector Search Compare output. Experimental MS3 reconstruction,
ternary crosslinks, adapters for other search engines, and a secondary
protein-pair evidence score remain on the project roadmap. In particular, the
current `URPsupport` feature materially improves large-dataset URP recovery but
should not be interpreted as a complete solution to protein-pair scoring.

A demonstration graphical interface is available at
[shinyapps.io](https://prospts.shinyapps.io/tstoneapp/), with additional
[instructions](https://msf.ucsf.edu/mike/crosslinkingClass/dataset_summary.html).

## References
